---
title: "Effect of Asymptomatic Transmission"
author: "Graham Casey Gibson"
date: "6/11/2020"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Effect of Asymptomatic Transmission on $R_t$ bias

We know from recent work that $R_t$ estimation is biased in the case of variable testing rates. However, it remains unclear how the sampling scheme effects the bias of the $R_t$ estimation. 


##### Experimental Setup

In order to evaluate the effect of asymptomatic carriers on $R_t$ bias we designed the following experimental setup.

We first simulate from a Susceptible-Infected-Asymptomatic-Recovered model (SIAR). 

for $S(t)$, $I(t)$,$A(t)$, and $R(t)$:
\begin{eqnarray}
\frac{dS}{dt} & = & - S(t)(\beta_1 I(t) + \beta_2 A(t))  \label{eq2a} \\
\frac{dI}{dt} & = & - S(t)\beta_1 I(t) - \gamma_1 I(t) \label{eq2b} \\
\frac{dA}{dt} & = & - S(t)\beta_2 A(t)- \gamma_2 A(t) \label{eq2c} \\
\frac{dR}{dt} & = & \gamma_1 I(t) + \gamma_2 A(t) . \label{eq2d}
\end{eqnarray}

and for initial conditions, 

$$I(0) =  N*(1-p)*1e-3$$
$$A(0) =  N*p*1e-3$$
$$R(0) = 0.0$$

$$S(0)= N- I(0) - A(0) -R(0)$$

where $N$ is the population size and $p$ is the percent of the infected people who are asymptomatic. This setup allows us to have both different $\beta$ and $\gamma$ depending on asymptomatic or symptomatic, reflecting the reality that asymptomatic carriers might recover more quickly and might more readily spread the disease due to lack of self-isolation. 

In order to estimate $R_t$ we use the `EpiEstim` package. This requires knowledge of the serial interval. However, for the compartmental model above, we know this is $(1-p)\frac{1}{\gamma_1} +p\frac{1}{\gamma_2}$.
##### Case 1

In the simplest case assume we assume that only symptomatic carriers are tested even though there are asymtpomatic people in the population. Furthermore, we suppose that both the asymptomatic and symptomatic carriers have the same transmission rate and the same recovery rate.  



```{r tabsets, echo=FALSE}
library(deSolve)
library(EpiEstim)
library(ggplot2)
run_analysis_at_percentage <- function(percentage,beta1,beta2){
      times      <- seq(0, 70, by = 1)
      print (beta1)
      parameters <- c(beta1 = beta1,beta2=beta2, gamma1 = .75,gamma2=.75)
      percent_asymptomatic  =  percentage
      init       <- c(S = 1-1e-3 ,A = percent_asymptomatic*1e-3, I_new= (1-percent_asymptomatic)*1e-3,I =(1-percent_asymptomatic)*1e-3, R = 0.0)
      
      ## Create an asymtpomatic SIR function
      sir_observe_symptoms <- function(time, state, parameters) {
        
        with(as.list(c(state, parameters)), {
          
          dS <- -  S * (beta1*I+ beta2*A )
          dA <- S*beta2*A  - gamma1*A
          dI <-   S * beta1*I - gamma2 * I
          dI_new <-    S * (beta1*I) 
          
          dR <- gamma1 * I + gamma2*A
          
          return(list(c(dS,dA,dI_new, dI, dR)))
        })
      }
      
      ### Set parameters
      ## Proportion in each compartment: Susceptible 0.999999, Infected 0.000001, Recovered 0
      ## beta: infection parameter; gamma: recovery parameter
      ## Time frame
      
      ## Solve using ode (General Solver for Ordinary Differential Equations)
      out_observe_symptoms <- ode(y = init, times = times, func = sir_observe_symptoms, parms = parameters)
      ## change to data frame
      out_observe_symptoms <- as.data.frame(out_observe_symptoms)
      ## Delete time variable
      ## multiply precentage to some population size
      out_observe_symptoms$I_new <- round(c(1e-6,diff(out_observe_symptoms$I_new))*1e8)
      
      ## Show data
      plot(out_observe_symptoms$I_new,type='l')
      
      
      ################################
      ## Create an asymtpomatic SIR function
      sir_asymptomatic <- function(time, state, parameters) {
        
        with(as.list(c(state, parameters)), {
          
          dS <- -  S * (beta1*I+ beta2*A )
          dA <- S*beta2*A  - gamma1*A
          dI <-   S * beta1*I - gamma2 * I
          dI_new <-    S * (beta1*I + beta2*A) 
          
          dR <- gamma1 * I + gamma2*A
          
          return(list(c(dS,dA,dI_new, dI, dR)))
        })
      }
      
      ### Set parameters
      
      ## Solve using ode (General Solver for Ordinary Differential Equations)
      out_asymptomatic <- ode(y = init, times = times, func = sir_asymptomatic, parms = parameters)
      ## change to data frame
      out_asymptomatic <- as.data.frame(out_asymptomatic)
      ## Delete time variable
      ## multiply precentage to some population size
      out_asymptomatic$I_new <- round(c(1e-6,diff(out_asymptomatic$I_new))*1e8)
      plot(out_asymptomatic$I_new,type='l')
      ## Show data
      
      r_t_hat <-estimate_R(out_asymptomatic$I_new,method=c("parametric_si"),
                           config = make_config(list(mean_si=1/parameters[3],
                                                     std_si=sqrt(1/parameters[3]))))

      r_t_hat_symptomatic <-estimate_R(out_observe_symptoms$I_new,method=c("parametric_si"),
                           config = make_config(list(mean_si=1/parameters[3],
                                                     std_si=sqrt(1/parameters[3]))))
      
      data_for_ggplot <- data.frame(t=seq(1,length((r_t_hat$R$`Mean(R)`))),
                                          r_t_symptomatic_only=r_t_hat_symptomatic$R$`Mean(R)`,
                                        r_t_true = r_t_hat$R$`Mean(R)`)
    
      p <- ggplot(data_for_ggplot,aes(x=t,y=r_t_true,col='True')) + geom_line(linetype = "dashed") +
           geom_line(aes(x=t,y=r_t_symptomatic_only,col='Only Observing Symptomatic'),linetype="dotted") + ylab("R_t") + xlab("t") + theme_bw()
      return (p)
      

}

library(shiny)

ui <- fluidPage(
  
  # App title ----

  # Sidebar layout with input and output definitions ----
  sidebarLayout(
    
    # Sidebar panel for inputs ----
    sidebarPanel(
      
      # Input: Slider for the number of bins ----
      numericInput(inputId = "pa",
                   label = "Percentage Asymptomatic",
                   value = 0),
      
      numericInput(inputId = "beta1",
                   label = "Symptomatic Transmission",
                   value = 1)
      
    ),
    
    # Main panel for displaying outputs ----
    mainPanel(
      
      # Output: Histogram ----
      plotOutput(outputId = "distPlot")
      
    )
  )
)
server <- function(input, output) {
  
  # Histogram of the Old Faithful Geyser Data ----
  # with requested number of bins
  # This expression that generates a histogram is wrapped in a call
  # to renderPlot to indicate that:
  #
  # 1. It is "reactive" and therefore should be automatically
  #    re-executed when inputs (input$bins) change
  # 2. Its output type is a plot
  output$distPlot <- renderPlot({
    
   
    
    
    p <- run_analysis_at_percentage(input$pa/100,input$beta1,input$beta1)
    print (p)
    
  })
  
}
shinyApp(ui, server)



```



Here we can see that no matter what values we put in for percent of asymptomatic people in the population, we get the same estimate for $R_t$, and that estimate has 0 bias. 



##### Case 2

In the second case assume we assume that only symptomatic carriers are tested even though there are asymtpomatic people in the population. However, we allow the asymptomatic transmission rates to differ from the symptomatic ones. 



```{r tabsets1, echo=FALSE}
library(deSolve)
library(EpiEstim)
library(ggplot2)
run_analysis_at_percentage <- function(percentage,beta1,beta2){
      times      <- seq(0, 70, by = 1)
      print (beta1)
      parameters <- c(beta1 = beta1,beta2=beta2, gamma1 = .75,gamma2=.75)
      percent_asymptomatic  =  percentage
      init       <- c(S = 1-1e-3 ,A = percent_asymptomatic*1e-3, I_new= (1-percent_asymptomatic)*1e-3,I =(1-percent_asymptomatic)*1e-3, R = 0.0)
      
      ## Create an asymtpomatic SIR function
      sir_observe_symptoms <- function(time, state, parameters) {
        
        with(as.list(c(state, parameters)), {
          
          dS <- -  S * (beta1*I+ beta2*A )
          dA <- S*beta2*A  - gamma1*A
          dI <-   S * beta1*I - gamma2 * I
          dI_new <-    S * (beta1*I) 
          
          dR <- gamma1 * I + gamma2*A
          
          return(list(c(dS,dA,dI_new, dI, dR)))
        })
      }
      
      ### Set parameters
      ## Proportion in each compartment: Susceptible 0.999999, Infected 0.000001, Recovered 0
      ## beta: infection parameter; gamma: recovery parameter
      ## Time frame
      
      ## Solve using ode (General Solver for Ordinary Differential Equations)
      out_observe_symptoms <- ode(y = init, times = times, func = sir_observe_symptoms, parms = parameters)
      ## change to data frame
      out_observe_symptoms <- as.data.frame(out_observe_symptoms)
      ## Delete time variable
      ## multiply precentage to some population size
      out_observe_symptoms$I_new <- round(c(1e-6,diff(out_observe_symptoms$I_new))*1e8)
      
      ## Show data
      plot(out_observe_symptoms$I_new,type='l')
      
      
      ################################
      ## Create an asymtpomatic SIR function
      sir_asymptomatic <- function(time, state, parameters) {
        
        with(as.list(c(state, parameters)), {
          
          dS <- -  S * (beta1*I+ beta2*A )
          dA <- S*beta2*A  - gamma1*A
          dI <-   S * beta1*I - gamma2 * I
          dI_new <-    S * (beta1*I + beta2*A) 
          
          dR <- gamma1 * I + gamma2*A
          
          return(list(c(dS,dA,dI_new, dI, dR)))
        })
      }
      
      ### Set parameters
      
      ## Solve using ode (General Solver for Ordinary Differential Equations)
      out_asymptomatic <- ode(y = init, times = times, func = sir_asymptomatic, parms = parameters)
      ## change to data frame
      out_asymptomatic <- as.data.frame(out_asymptomatic)
      ## Delete time variable
      ## multiply precentage to some population size
      out_asymptomatic$I_new <- round(c(1e-6,diff(out_asymptomatic$I_new))*1e8)
      plot(out_asymptomatic$I_new,type='l')
      ## Show data
      
      r_t_hat <-estimate_R(out_asymptomatic$I_new,method=c("parametric_si"),
                           config = make_config(list(mean_si=1/parameters[3],
                                                     std_si=sqrt(1/parameters[3]))))

      r_t_hat_symptomatic <-estimate_R(out_observe_symptoms$I_new,method=c("parametric_si"),
                           config = make_config(list(mean_si=1/parameters[3],
                                                     std_si=sqrt(1/parameters[3]))))
      
      data_for_ggplot <- data.frame(t=seq(1,length((r_t_hat$R$`Mean(R)`))),
                                          r_t_symptomatic_only=r_t_hat_symptomatic$R$`Mean(R)`,
                                        r_t_true = r_t_hat$R$`Mean(R)`)
    
      p <- ggplot(data_for_ggplot,aes(x=t,y=r_t_true,col='True')) + geom_line(linetype = "dashed") +
           geom_line(aes(x=t,y=r_t_symptomatic_only,col='Only Observing Symptomatic'),linetype="dotted") + ylab("R_t") + xlab("t") + theme_bw()
      return (p)
      

}

library(shiny)

ui <- fluidPage(
  
  # App title ----

  # Sidebar layout with input and output definitions ----
  sidebarLayout(
    
    # Sidebar panel for inputs ----
    sidebarPanel(
      
      # Input: Slider for the number of bins ----
      numericInput(inputId = "pa",
                   label = "Percentage Asymptomatic",
                   value = .2),
      
      numericInput(inputId = "beta1",
                   label = "Symptomatic Transmission",
                   value = 1),
      numericInput(inputId = "beta2",
                   label = "Asymptomatic Transmission",
                   value = 2 )
      
      
    ),
    
    # Main panel for displaying outputs ----
    mainPanel(
      
      # Output: Histogram ----
      plotOutput(outputId = "distPlot")
      
    )
  )
)
server <- function(input, output) {
  
  # Histogram of the Old Faithful Geyser Data ----
  # with requested number of bins
  # This expression that generates a histogram is wrapped in a call
  # to renderPlot to indicate that:
  #
  # 1. It is "reactive" and therefore should be automatically
  #    re-executed when inputs (input$bins) change
  # 2. Its output type is a plot
  output$distPlot <- renderPlot({
    
   
    
    
    p <- run_analysis_at_percentage(input$pa/100,input$beta1,input$beta2)
    print (p)
    
  })
  
}
shinyApp(ui, server)



```


 Here we can see that if the asymptomatic transmission rate is higher than the symtpomatic transmission, we bias the estiamte of $R_t$. This magnitude of the bias now depends on the parameter $p$, with more asymtpomatic cases producing more bias. 
 
 
##### Case 3

In the third case assume we assume that only symptomatic carriers are tested even though there are asymtpomatic people in the population. However, we allow the asymptomatic transmission rates to differ from the symptomatic ones, while also setting the recovery rates so that asymptomatic people recover twice as fast as symptomatic people. 



```{r tabsets2, echo=FALSE}
library(deSolve)
library(EpiEstim)
library(ggplot2)
run_analysis_at_percentage <- function(percentage,beta1,beta2){
      times      <- seq(0, 70, by = 1)
      print (beta1)
      parameters <- c(beta1 = beta1,beta2=beta2, gamma1 = .75,gamma2=.75/6)
      percent_asymptomatic  =  percentage
      init       <- c(S = 1-1e-3 ,A = percent_asymptomatic*1e-3, I_new= (1-percent_asymptomatic)*1e-3,I =(1-percent_asymptomatic)*1e-3, R = 0.0)
      
      ## Create an asymtpomatic SIR function
      sir_observe_symptoms <- function(time, state, parameters) {
        
        with(as.list(c(state, parameters)), {
          
          dS <- -  S * (beta1*I+ beta2*A )
          dA <- S*beta2*A  - gamma1*A
          dI <-   S * beta1*I - gamma2 * I
          dI_new <-    S * (beta1*I) 
          
          dR <- gamma1 * I + gamma2*A
          
          return(list(c(dS,dA,dI_new, dI, dR)))
        })
      }
      
      ### Set parameters
      ## Proportion in each compartment: Susceptible 0.999999, Infected 0.000001, Recovered 0
      ## beta: infection parameter; gamma: recovery parameter
      ## Time frame
      
      ## Solve using ode (General Solver for Ordinary Differential Equations)
      out_observe_symptoms <- ode(y = init, times = times, func = sir_observe_symptoms, parms = parameters)
      ## change to data frame
      out_observe_symptoms <- as.data.frame(out_observe_symptoms)
      ## Delete time variable
      ## multiply precentage to some population size
      out_observe_symptoms$I_new <- round(c(1e-6,diff(out_observe_symptoms$I_new))*1e8)
      
      ## Show data
      plot(out_observe_symptoms$I_new,type='l')
      
      
      ################################
      ## Create an asymtpomatic SIR function
      sir_asymptomatic <- function(time, state, parameters) {
        
        with(as.list(c(state, parameters)), {
          
          dS <- -  S * (beta1*I+ beta2*A )
          dA <- S*beta2*A  - gamma1*A
          dI <-   S * beta1*I - gamma2 * I
          dI_new <-    S * (beta1*I + beta2*A) 
          
          dR <- gamma1 * I + gamma2*A
          
          return(list(c(dS,dA,dI_new, dI, dR)))
        })
      }
      
      ### Set parameters
      
      ## Solve using ode (General Solver for Ordinary Differential Equations)
      out_asymptomatic <- ode(y = init, times = times, func = sir_asymptomatic, parms = parameters)
      ## change to data frame
      out_asymptomatic <- as.data.frame(out_asymptomatic)
      ## Delete time variable
      ## multiply precentage to some population size
      out_asymptomatic$I_new <- round(c(1e-6,diff(out_asymptomatic$I_new))*1e8)
      plot(out_asymptomatic$I_new,type='l')
      ## Show data
      
      r_t_hat <-estimate_R(out_asymptomatic$I_new,method=c("parametric_si"),
                           config = make_config(list(mean_si=1/parameters[3],
                                                     std_si=sqrt(1/parameters[3]))))

      r_t_hat_symptomatic <-estimate_R(out_observe_symptoms$I_new,method=c("parametric_si"),
                           config = make_config(list(mean_si=1/parameters[3],
                                                     std_si=sqrt(1/parameters[3]))))
      
      data_for_ggplot <- data.frame(t=seq(1,length((r_t_hat$R$`Mean(R)`))),
                                          r_t_symptomatic_only=r_t_hat_symptomatic$R$`Mean(R)`,
                                        r_t_true = r_t_hat$R$`Mean(R)`)
    
      p <- ggplot(data_for_ggplot,aes(x=t,y=r_t_true,col='True')) + geom_line(linetype = "dashed") +
           geom_line(aes(x=t,y=r_t_symptomatic_only,col='Only Observing Symptomatic'),linetype="dotted") + ylab("R_t") + xlab("t") + theme_bw()
      return (p)
      

}

library(shiny)

ui <- fluidPage(
  
  # App title ----

  # Sidebar layout with input and output definitions ----
  sidebarLayout(
    
    # Sidebar panel for inputs ----
    sidebarPanel(
      
      # Input: Slider for the number of bins ----
      numericInput(inputId = "pa",
                   label = "Percentage Asymptomatic",
                   value = .2),
      
      numericInput(inputId = "beta1",
                   label = "Symptomatic Transmission",
                   value = 1),
      numericInput(inputId = "beta2",
                   label = "Asymptomatic Transmission",
                   value = 1)
      
      
    ),
    
    # Main panel for displaying outputs ----
    mainPanel(
      
      # Output: Histogram ----
      plotOutput(outputId = "distPlot")
      
    )
  )
)
server <- function(input, output) {
  
  # Histogram of the Old Faithful Geyser Data ----
  # with requested number of bins
  # This expression that generates a histogram is wrapped in a call
  # to renderPlot to indicate that:
  #
  # 1. It is "reactive" and therefore should be automatically
  #    re-executed when inputs (input$bins) change
  # 2. Its output type is a plot
  output$distPlot <- renderPlot({
    
   
    
    
    p <- run_analysis_at_percentage(input$pa/100,input$beta1,input$beta2)
    print (p)
    
  })
  
}
shinyApp(ui, server)



```


Here we see no bias introduced by the different recovery times